Variance reduction for Markov chains via martingale representations

### 2020-04-14 aap/Revision/vr-mcmc-revision-new-new.tex
update vr-mcmc-revision-new-new.tex

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
\begin{eqnarray}
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-(1/2)(x-\mu)^T\Sigma^{-1}(x-\mu)} + \rme^{-(1/2)(x+\mu)^T\Sigma^{-1}(x+\mu)}\right)
\end{eqnarray}
where $\rho \in (0,1)$, $\mu \in \rset^2$, $\Sigma$ is a symmetric positive definite $2 \times 2$ matrix and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\]
and
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. Hence we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
\par

@@

\subsection{Banana shape distribution}
The “Banana-shape” distribution, proposed by \cite{haario1999adaptive}, can be obtained from a d-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
\[
\varphi(x_1,\ldots,x_n) = (x_1, x_2 + bx^2_1 - pb,x_3,\ldots,x_d)
\]
where $p > 0$ and $b > 0$ are parameters; $b$ accounts for the curvature of density’s level sets. The potential $U$ is given by
\[
U(x_1,\ldots, x_d) = \frac{x_1^2}{2} + \frac{(x_2 + bx_1^2 - pb)^2}{2} + \sum\limits_{k=3}^{d}\frac{x_k^2}{2}
\]
The quantity of interest is the expectation of $f(x) = x_2$. We set $p = 100, b = 0.1$ and consider $d=8$. We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$. We use $T= 5 \times 10^4$ independent training trajectories, each of size $n=50$ with the burn-in $n_{\text{burn}} = 100$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.

@@

\subsection{Binary Logistic Regression}
Second experiment considers the problem of logistic regression. Let $\mathsf{Y} = \left(\mathsf{Y}_1,\ldots,\mathsf{Y}_n\right) \in \{0,1\}^m$ be binary response variables, $\mathsf{X} \in \rset^{m \times d}$ be a feature matrix and $\theta \in \rset^d$ - vector of regression parameters. We define log-likelihood of $i$-th observation as
\[
\ell(\mathsf{Y}_i | \theta, \mathsf{X}_i) = \mathsf{Y}_i \mathsf{X}_i^T\theta - \ln{\left(1 + \rme^{\mathsf{X}_i^T\theta}\right)}
\]
In order to estimate $\theta$ according to given data, the Bayesian approach introduces prior distribution $\pi_0(\theta)$ and inferences the posterior density $\pi(\theta| \mathsf{Y}, \mathsf{X})$ using Bayes' rule.

@@

\section{Proofs}
\label{sec:proofs}
\subsection{Proof of Theorem~\ref{thm:main-repr}}
The expansion obviously holds for $p = q=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
The orthonormality and completeness
of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$,
$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
\[
f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
\]
Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
\begin{eqnarray*}
\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left. \PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left. \PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
\\
&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
\end{eqnarray*}
Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
\begin{equation}
\label{eq:1-step-decomp}
f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
\end{equation}
where
\[
a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
\]
which is the required statement in the case $q = m+1$ and $j=m$.
Now assume that $j<m$.
Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$. Note that $\mathsf{P}$-a.s. it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction. Hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
 = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
\end{align*}
where for $y \in \rset^d$,
\[
a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
\]
Formulas \eqref{eq:1-step-decomp}
and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.

\subsection{Proof of Lemma~\ref{lem:variance}}
Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in Lemma~\ref{lem:v_ergodicity}. Due to lemma~\ref{lem:covariance},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
\]
and the result of lemma follows from Taylor expansion of denominator.

@@

\begin{proof}
Let $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$,
$z\in\mathbb R^d$, denote the density of a $d$-dimensional
standard Gaussian random vector.
We first remark that, for the normalized Hermite polynomial $\mathbf H_{\mathbf k}$ on $\mathbb R^d$,
$\mathbf k\in\mathbb N_0^d$, it holds
$$
\mathbf{H}_{\mathbf{k}}(z)\boldsymbol{\varphi}(z)
=\frac{(-1)^{|\mathbf{k}|}}{\sqrt{\mathbf{k} !}} \partial^{\mathbf{k}} \boldsymbol{\varphi}(z).
$$
This enables to use the integration by parts in vector form as follows
(below $\prod_{j=l+1}^p:=1$ whenever $l=p$)
\begin{align*}
&\bar a_{p,\mathbf{k}}(x)
 =
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
\mathbf{H}_{\mathbf{k}}(z_{1})\boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
\\
& =
\frac{1}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots  \rmd z_{p}
\\
& =
\frac{\gamma^{|\mathbf k'|/2}}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
\partial_{1}^{\mathbf k'}[f\circ G_{p}](x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
\\
 & =\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[\partial_{y_{1}}^{\mathbf{k}'}[f\circ G_{p}](x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right].
\end{align*}
The last expression yields the result.
\end{proof}

@@

Note that conditions {\bf (H1)} and {\bf (H2)} imply (\cite[Lemma~16]{brosse2018diffusion}) that for some $K_2 > 0$ it holds
\begin{equation}
\label{eq:H3}
\ps{\nabla U(x)}{x} \geq \frac{m}{2}\|x\|^2, \quad \|x\| \geq K_2
\end{equation}

Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the drift condition \eqref{eq:lyapunov}

\begin{lem}
\label{lem:drift}
 Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$. 
 Then the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition \eqref{eq:lyapunov} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L}$ with drift function $V(x) = 1 + \|x\|^2$,   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_2^2}{8} + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
\end{lem}

\begin{proof} Let $V(x) = 1 + \|x\|^2$, then
$$
Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}\frac{(1+\|y\|^2)}{(2\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{2\gamma}\right)}\,dy =
$$

>>>>

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
\begin{eqnarray}
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-(1/2)(x-\mu)^T\Sigma^{-1}(x-\mu)} + \rme^{-(1/2)(x+\mu)^T\Sigma^{-1}(x+\mu)}\right)
\end{eqnarray}
where  $\mu \in \rset^d$, $\Sigma$ is a positive-definite $d \times d$ matrix and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\]
and
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2 \left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1} \Sigma^{-1}\mu ,
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 5 \times 10^4$ independent training trajectories, each one of size $n = 50$ with the burn-in period $n_{\text{burn}} = 100$. Then we solve the least squares  problems \eqref{eq:Qregr} using the first order polynomial approximations for $f(x)=\sum_{i=1}^d x_i$ and second order approximations for $f(x) = \sum_{i=1}^d x^2_i$ as described in the previous section. Hence we set $K = 1$ for $f(x)=\sum_{i=1}^d x_i$ or $K = 2$ for $f(x) = \sum_{i=1}^d x^2_i$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and plot boxplots of the corresponding ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials in case of $K=1$ and second order polynomials for $K = 2$.
\par

@@

\subsection{Banana shape distribution}
The “Banana-shape” distribution, proposed by \cite{haario1999adaptive}, can be obtained from a $d$-dimensional Gaussian vector with zero mean and covariance $\mathrm{diag}(p,1,\ldots,1)$ by applying transformation $\varphi_b(x) : \rset^d \rightarrow \rset^d$ of the form
\[
\varphi_b(x_1,\ldots,x_n) = (x_1, x_2 + bx^2_1 - pb,x_3,\ldots,x_d)
\]
where $p > 0$ and $b > 0$ are parameters; $b$ accounts for the curvature of density’s level sets. The potential $U$ is given by
\[
U(x_1,\ldots, x_d) = \frac{x_1^2}{2} + \frac{(x_2 + bx_1^2 - pb)^2}{2} + \frac{1}{2} \sum_{k=3}^{d} x_k^2 \eqsp.
\]
The quantity of interest is the expectation of $f(x) = x_2$. We set $p = 100, b = 0.1$ and consider $d=8$. We solve the least squares  problems \eqref{eq:Qregr} using the second-order polynomial approximations for the coefficients \(\hat{a}_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 2$. We use $T= 5 \times 10^4$ independent training trajectories, each of size $n=50$ with the burn-in $n_{\text{burn}} = 100$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 2 \times 10^3$ and display boxplots of the ergodic averages in Figure~\ref{fig:banana_shape}.

@@

\subsection{Binary Logistic Regression}
In this section, we consider  logistic regression. Let $\mathsf{Y} = \left(\mathsf{Y}_1,\ldots,\mathsf{Y}_n\right) \in \{0,1\}^m$ be binary response variables, $\mathsf{X} \in \rset^{m \times d}$ be a feature matrix and $\theta \in \rset^d$ - vector of regression parameters. We define log-likelihood of $i$-th observation as
\[
\ell(\mathsf{Y}_i | \theta, \mathsf{X}_i) = \mathsf{Y}_i \mathsf{X}_i^T\theta - \ln{(1 + \rme^{\mathsf{X}_i^T\theta})}
\]
In order to estimate $\theta$ according to given data, the Bayesian approach introduces prior distribution $\pi_0(\theta)$ and consider the posterior density $\pi(\theta| \mathsf{Y}, \mathsf{X})$ using Bayes' rule.

@@

\section{Proofs}
\label{sec:proofs}

\subsection{Proof of Theorem~\ref{thm:main-repr}}
The expansion obviously holds for $p = q=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
The orthonormality and completeness
of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$,
$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
\[
f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
\]
Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
\begin{eqnarray*}
\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left. \PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left. \PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
\\
&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
\end{eqnarray*}
Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
\begin{equation}
\label{eq:1-step-decomp}
f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
\end{equation}
where
\[
a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
\]
which is the required statement in the case $q = m+1$ and $j=m$.
Now assume that $j<m$.
Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$. Note that $\mathsf{P}$-a.s. it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction. Hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
 = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
\end{align*}
where for $y \in \rset^d$,
\[
a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
\]
Formulas \eqref{eq:1-step-decomp}
and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.

\subsection{Proof of Lemma~\ref{lem:variance}}
In the sequel, the notation $A(\gamma,n,x) \lesssim B(\gamma,n,x)$ means that there exist $\gamma_0 > 0$, and $C < \infty$ such that for all  $\gamma \in (0,\gamma_0]$, $x \in \rset^d$, $n \in \nset$, $A(\gamma,n,x) \leq C B(\gamma,n,x)$.

Under assumptions {\bf (H1)} and {\bf (H2)} the Markov chain $(X^x_{p})_{p \in \nset_0}$ is $V-$geometrically ergodic with $V(x) = 1+\|x\|^2$ and $\rho = \rme^{-\kappa\gamma}$ with $\kappa$ specified in Lemma~\ref{lem:v_ergodicity}. Due to lemma~\ref{lem:covariance},
\[
\PVar[\pi_n^N(f)]\lesssim \frac{1}{n(1-\rho^{1/2})}
\]
and the result of lemma follows from Taylor expansion of denominator.

@@

\begin{proof}
Note that for the normalized Hermite polynomial $\mathbf H_{\mathbf k}$ on $\mathbb R^d$,
$\mathbf k\in\mathbb N_0^d$, it holds
$$
\mathbf{H}_{\mathbf{k}}(z)\boldsymbol{\varphi}(z)
=\frac{(-1)^{|\mathbf{k}|}}{\sqrt{\mathbf{k} !}} \partial^{\mathbf{k}} \boldsymbol{\varphi}(z).
$$
This enables to use the integration by parts in vector form as follows
(below $\prod_{j=l+1}^p:=1$ whenever $l=p$)
\begin{align*}
&\bar a_{p,\mathbf{k}}(x)
 =
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
\mathbf{H}_{\mathbf{k}}(z_{1})\boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
\\
& =
\frac{1}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots  \rmd z_{p}
\\
& =
\frac{\gamma^{|\mathbf k'|/2}}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
\partial_{1}^{\mathbf k'}[f\circ G_{p}](x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
\\
 & =\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[\partial_{y_{1}}^{\mathbf{k}'}[f\circ G_{p}](x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right].
\end{align*}
The last expression yields the result.
\end{proof}

@@

In all the sequel, without loss of generality, we assume that $\nabla U(0)=0$ and we set
\begin{equation}
\label{eq:definition-V}
V(x) = 1 + \|x\|^2 \eqsp.
\end{equation}
\begin{lem}
  \label{lem:quadratic_behaviour}
  Assume {\bf (H1)} and {\bf (H2)}. Then there exists $K_2 \geq 0$ such that for any $ \| x \| \geq K_2$ $\ps{\nabla U(x)}{x} \geq (m/2) \| x \|^2$.
\end{lem}
\begin{proof}
  Using {\bf (H1)} and {\bf (H2)}, we have for any $x \in \rset^d$, $\| x \|\geq K_1$,
  \begin{align*}
    \ps{\nabla U(x)}{x}
    &= \int_{0}^{K_1/\| x \|} D^2 U(t x ) [x^{\otimes 2}] \rmd t + \int_{K_1/\| x \|} ^ 1 D^2 U(t x ) [x^{\otimes 2}] \rmd t\\
    & \geq m \|x \|^2 \{1- K_1 (1 +L/m)   / \| x \| \}\eqsp,
  \end{align*}
which proves the first statement. The second statement is obvious.
\end{proof}
\begin{lem}
\label{lem:drift}
 Assume that {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$.
 Then, for any $\gamma \in (0,\bar{\gamma}]$ with $\overline{\gamma} = m/4$  the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition
 \begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
\end{equation}
with   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = 25K_2^2/8 + d + m$ with $K_2$ from \eqref{eq:H3}, $m$ from {\bf (H2)}.
\end{lem}

\begin{proof} 
$$
Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}\frac{(1+\|y\|^2)}{(2\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{2\gamma}\right)}\,dy =
$$
